An exposition on the propriety of restricted Boltzmann machines

### 2016-09-17 writing/draft.Rmd
add effective sample size test and nordman's proofs

This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}. Additionally, there are two accompanying appendices with auxiliary proofs relating to the choice of data encoding and parameter behavior during fitting with incomplete training data in section \ref{appendix}.

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$. The proof of this is included in appendix \ref{center-of-the-universe}. And while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

However, computational concerns are not the only issues with fitting an RBM using ML. In addition, the RBM model, with the appropriate choice of parameters, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the -1/1 valued data outcomes). This hyper-flexibility dictates that ML fitting will merely replicate the empirical data distribution, which is the optimal nonparametric solution. In a data application with a large sample space it is unlikely that the training set will include one of each possible image, thus there will be possible images that are not realized in the training set, unlike this small example. The presence of empty cells in the empirical distribution drives coordinates of $\boldsymbol \theta$ to $\infty$ in order to make the corresponding cell probabilities arbitrarily close to zero. As we have already discussed in secion \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM. A detailed discussion of this phenomenon are included in appendix \ref{incomplete-cases}.

@@

#Appendix

##Center of the universe

**TODO: Add**

##Incomplete cases

**TODO: Add**

>>>>

This paper is structured as follows, section \ref{background} gives a definition of the RBM including the joint distribution of hidden and visible variables and shows the connection to deep learning. Additionally, measures of model impropriety and methods of detecting them are defined. Section \ref{explorations} details our explorations into the model behavior and propriety (or lack thereof) of the RBM class. We discuss three Bayesian fitting techniques with the goal of avoiding model impropriety in section \ref{bayesian-model-fitting} and wrap up with a discussion in section \ref{discussion}. Additionally, there is an accompanying appendix with auxiliary proofs relating to the choice of data encoding in section \ref{appendix}.

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{H + V + H*V}$. The proof of this is included in appendix \ref{on-the-one-to-one-correspondence-between parameters-and moments-in-the-rbm-distribution}. And while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

However, computational concerns are not the only issues with fitting an RBM using ML. In addition, the RBM model, with the appropriate choice of parameters, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the -1/1 valued data outcomes). To see this, we provide a small example.

For simplicity, consider a model with two visible variables $(V_1,V_2)$ and one hidden $H_1$.  In this case,
there are four possible data realizations for $(V_1,V_2)$  given by $(\pm 1, \pm 1)$ and we may express the model probabilities
as
$$
P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
$$
for $(v_1,v_2)\in \{-1, 1\}^2$, in terms of real-valued parameters $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}$. Given any  specified cell probabilities, say
\begin{equation}
\label{eqn:0}
0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
\end{equation} 
for the outcomes $(\pm 1, \pm 1)$,  values for parameters $(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may be chosen to mimic such  cell probabilities with arbitrary closeness.  In fact, when the cell probabilities (\ref{eqn:0}) are all strictly positive, parameters in the RBM model can be specified to reflect these probabilities exactly. And, when one or more of the cell probabilities (\ref{eqn:0}) are zero, the corresponding RBM  probabilities $P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})$ may never be identically zero (due to the exponential model form) but
 parameters can be still selected to make the appropriate RBM cell probabilities arbitrarily small.

To demonstrate such model approximations,  we  assume $p_{(-1,-1)}>0$ (without loss of generality) in the specified cell probabilities (\ref{eqn:0}) and replace parameters $\theta_{11},\theta_{21}$ with $\Delta_1 \equiv \theta_{11} +\theta_{21}$ and $\Delta_2 \equiv \theta_{11} -\theta_{21}$.  We may then prescribe values of $\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2$  so that the model probability ratio
$$
P(V_1=v_1,V_2=v_2|\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)/  P(V_1=-1,V_2=-1| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \Delta_1,\Delta_2)
$$
matches the corresponding  ratio $p_{(v_1,v_2)}/p_{(-1,-1)}$  over three values of $(v_1,v_2) = (1,-1),(-1,1),(1,1)$.

For instance, assuming the cell probabilities from (\ref{eqn:0}) are all positive, these probabilities
 can be exactly reproduced by choosing
 \begin{eqnarray*}
 \theta_{v_1} &=& \frac{1}{2} \log\left( \frac{p_{(1,-1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}+\Delta_2) + \exp(-\theta_{h_1}-\Delta_2)}  \right),\\ 
 \theta_{v_2}&= &\frac{1}{2} \log\left( \frac{p_{(-1,1)}}{p_{(-1,-1)}} \frac{\exp(\theta_{h_1}-\Delta_1) + \exp(-\theta_{h_1}+\Delta_1)}{\exp(\theta_{h_1}-\Delta_2) + \exp(-\theta_{h_1}+\Delta_2)}  \right)
\end{eqnarray*}
and selecting $\theta_{h_1}, \Delta_1,\Delta_2$ to solve
\begin{equation}
\label{eqn:1}
\frac{p_{(1,1)}p_{(-1,-1)}}{p_{(-1,1)}p_{(1,-1)}} = \frac{\ell(|\theta_{h_1}|) + \ell(|\Delta_1|) }{\ell(|\theta_{h_1}|) + \ell(|\Delta_2|)},
\end{equation}
based on a monotonically increasing function $\ell(x)\equiv \exp(-2x)+\exp(2x)$, $x \geq 0$. If  $[p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] \geq 1$, one can pick any values for $\theta_{h_1}, \Delta_2\in \mathbb{R}$ and solve (\ref{eqn:1}) for $|\Delta_1|$; likewise, when $[p_{(1,1)}p_{(-1,-1)}]/ [p_{(-1,1)}p_{(1,-1)}] < 1$ in (\ref{eqn:1}), one may solve for $|\Delta_2|$ upon choosing any values for $\theta_{h_1}, \Delta_1\in \mathbb{R}$.

Alternatively, if exactly one specified cell probability in (\ref{eqn:0}) is zero, say $p_{(1,1)}$ (without loss of generality), we can select parameters $\theta_{v_1},\theta_{v_2}$ as above based on a sequence $(\theta_{h_1}, \Delta_{1}, \Delta_{2}) \equiv (\theta_{h_1}^{(m)}, \Delta_{1}^{(m)}, \Delta_{2}^{(m)})$, $m\in\{1,2,\ldots,\}$ of the remaining parameter values such that $\lim_{m\to \infty}|\Delta_{1}^{(m)}| = \infty$ and $\lim_{m\to \infty}  (|\theta_{h_1}^{(m)}| + |\Delta_{2}^{(m)}|)/|\Delta_{1}^{(m)}|=0$ hold. This guarantees that the resulting RBM model matches the given cell probabilities (\ref{eqn:0}) in the limit:
\begin{equation}
\label{eqn:2}
\lim_{m\to \infty}P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1},\Delta_1,\Delta_2) = p_{(v_1,v_2)},\quad (v_1,v_2)\in\{(\pm 1,\pm 1)\}.
\end{equation}  
If exactly two specified probabilities in (\ref{eqn:0}) are zero, say  $p_{(1,1)}$ and $p_{(-1,1)}$ (without loss of generality), then a limit approximation as in (\ref{eqn:2}) follows by picking $\theta_{v_1}$ as above based on any choices of $(\theta_{h_1}, \Delta_1,\Delta_2)$ and choosing  a sequence of $\theta_{v_2}\equiv \theta_{v_2}^{(m)}$ values for which $\theta_{v_2}^{({m})} \rightarrow -\infty$.

The previous discussion intends to indicate that the RBM model suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization. Quite different parameter settings can induce the same essential RBM model.However, this  is not the most disastrous implication of the RBM parameterization. A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know what the answer must be before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the data. That is, based on a random sample of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well approximated.

For illustration, continue the simple example from above with $n$ iid observations, each consisting of two realized visibles $(V_1,V_2)$.  In which case, when the specified cell probabilities $p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}$ in (\ref{eqn:0}) are taken as the empirical cell frequencies from the sample, there is no better model based on maximum likelihood and the discussion above (c.f. (\ref{eqn:2})) shows that RBM model parameters can be chosen to re-create this empirical distribution to an arbitrarily close degree. Hence, RBM model fitting based on ML will often simply seek to reproduce the empirical distribution (i.e., as an extreme type of over-fitting). What's more, whenever this empirical distribution contains empty cells, fitting steps for the RBM model will necessarily aim to chase parameters that necessarily diverge in magnitude in order to zero out the corresponding RBM cell probabilities. In a data application with a large sample space it is unlikely that the training set will include one of each possible image, thus there will be possible images that are not realized in the training set, unlike this small example. As we have already discussed in secion \ref{explorations}, large magnitudes of $\boldsymbol \theta$ lead to model impropriety in the RBM.

@@

# Appendix

## On the one-to-one correspondence between parameters and moments in the RBM distribution

```{r child='appendix.Rmd'}
```
